---
title: "CalgaryVoting"
author: "L.Doyle"
date: "07/03/2021"
output: 
  html_document: 
  keep_md: true
 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calgary Voting and Demographics  
This is a work in progress and very much still in it's early stages.  

This is an exploratory analysis to explore relationshsips between demographics of a community and their voting tendencies.  I will begin by looking for patterns in age and gender distribution compared to election results from the 2017 Civic election.  After that I may look to add additional data like provincial and federal voting habits, housing income, single family versus multi unit housing, historical civic voting habits, advance voters trends or voters from seniors residence, etc.  Data was downloaded from the City of Calgary's Open Data website. 

### Assumptions and uncertainties
Taking community demographic data and comparing it to election results from that community will not be a perfect match since multiple communities will feed into a single polling station and certain demographics are likely over-represented in voters but it will make a good first approximation.I would also like to do some statistical analysis on any perceived trends and relationships to see if they are significant.

```{r loading, warning = FALSE, message = FALSE}
library(tidyverse)
# load data
Elect <- read.csv("./2017_Official_Election_Results_by_Voting_Station.csv")
ComDemo <- read.csv("./Civic_Census_by_Community__Age_and_Gender.csv")
```

### Civic Census by Community, Age, Gender from 1996 to 2019  
[link](https://data.calgary.ca/Demographics/Civic-Census-by-Community-Age-and-Gender/vsk6-ghca)
```{r census}
GenderReport <- ComDemo %>%
  group_by(YEAR) %>%
  summarize(males = sum(MALES), females = sum(FEMALES)) %>%
  print()
```

The above table indicates that there was no gender reporting in the 1996 Civic Census, filter out 1996.

```{r ElectClean2}
ComDemo <- ComDemo %>% filter(YEAR != "1996")
```

### 2017 Calgary Civic Election Results.  
[link](https://data.calgary.ca/Government/2017-Official-Election-Results-by-Voting-Station/atsy-3a4w).  

#### Election Data Cleaning  
Election results include school board trustees we only care about city councillors and Mayor for now also remove unwanted columns.
```{r ElectClean}
Elect <- Elect %>% 
  filter(Office %in% c("COUNCILLOR", "MAYOR")) %>%
  select(-Voting.Station.ID)
```

#### Voting Station Communities  
Election results are reported by voting station but do not have communities associated with the voting stations.  Demographic data is reported by community. To join these two datasets I need to get the communities that the voting stations are in. I did this by loading "voting station location" csv [link](https://data.calgary.ca/Government/Voting-Stations-Effective-October-16-2017-/ps5q-maip) and "community boundaries" shapefile [link](https://data.calgary.ca/Base-Maps/Community-Boundaries/surr-xmvs) (downloaded from the city of Calgary) into QGIS and creating an intersection layer which includes a community column in the voting stations table. I then exported that layer and imported it into R. 
```{r StationsWCommunity}
stationsWcomm <- read.csv("./VotingStationsWCommunity.csv")

# Get rid of unwanted columns
stationsWcomm <- stationsWcomm %>%
  select(VSTN_ID, NAME, comm_code, name_2)

# Join stationsWcomm df w the Elect df, match by name rather than stationID as some stations have multiple ID
ElectComm <- Elect %>% left_join(stationsWcomm, by = c("Voting.Station.Name" = "NAME"))

# investigate the voting stations that don't correspond to a community & how many votes are associated w them
NoCommunity <- subset(ElectComm, is.na(comm_code))
sum(NoCommunity$Votes) # votes not associated w a community
sum(Elect$Votes) # total votes;
```
22694/752708  = 3.0%.  3% of votes will not be accounted for if I remove these rows.  

#### How are the unaccounted for votes classified
Determine what type of voting stations the votes with no associated community are from.
```{r notsure}
table(NoCommunity$Voting.Station.Type)  
```


## Votes not assigned to a voting station  
Try to make some prediction about the demographic of the unassigned votes and investigate to make sure there is nothing unusual about the votes before filtering them from the dataset

```{r unaccounted}
NoCommunityType <- NoCommunity %>%
  group_by(Voting.Station.Type) %>%
  summarize(Votes2 = sum(Votes))
            
NoCommunityType <- mutate(NoCommunityType, percent = round(Votes2/sum(NoCommunityType$Votes2)*100, 0))
NoCommunityType
```
This table indicates that 23% of the NoCommunity votes are from a "special" Voting.Station.Type which appears to be the designation given for seniors residence. These votes could be reasonably assigned to +65 demographic.
**I need to do something with the below table to compare it to overall voting habits and make sure it all checks out. Maybe create visuals for each ward comparing unclassified votes to in-person votes to look for descrepencies**
```{r discrepencies}
# Investigate there is nothing unusual about these votes
UnaccountedVotes <- NoCommunity %>%
  group_by(Office, Ward, Ballot.Name) %>%
  summarize(Votes2 = sum(Votes)) %>%
  arrange(desc(Votes2), .by_group = TRUE) %>%
  print()

## filter out the votes that are not assigned to a community
CommunityVotes <- subset(ElectComm, !is.na(comm_code))
```
## Ward 11
I started by looking at distribution of votes for Ward 11 City Councillor. Analysis indicates that 20% of Ward11 votes were cast in advance and are thus not associated with a voting station. I will only be looking at votes associated with a voting station so that I can compare them with community demographics in phase two. **I should come back and look at the vote distribution in advance voters to see if it matches in-person voting trends if possible.**
```{r Ward11, fig.width =10}
# create a df of just Ward11 
Ward11 <- filter(CommunityVotes, Ward == "11", Office == "COUNCILLOR") 

# Determine percentage of votes cast in advance in Ward11
Ward11Advance <- Ward11 %>%
  group_by(Voting.Station.Type) %>%
  summarize(Votes2 = sum(Votes))
Ward11Advance <- mutate(Ward11Advance, percent = round(Votes2/sum(Votes2)*100,0))
Ward11Advance

# Looking only at regular in-person voting
Ward11Regular <- Ward11 %>%
  filter(Voting.Station.Type == "Regular") %>%
  arrange(Voting.Station)
Ward11Percents <- Ward11Regular %>%
  group_by(Voting.Station) %>%
  mutate(percent = round(Votes/sum(Votes)*100, 1)) %>%
  relocate(Ward, Ballot.Name, Votes, percent, Voting.Station.Name, comm_code, name_2) ## rearranges columns
Ward11Councillors <- Ward11Percents %>%
  arrange(Ballot.Name)
Ward11Councillors

# plot the percentage of vote for each candidate grouped by polling station
gWard11Col <- ggplot(Ward11Councillors, aes(fill = Ballot.Name, y = percent, x = name_2)) +
  geom_col(position = position_dodge(0.5), width = 0.5) +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(x="Polling Station Community", y = "Percent of Voting Station Vote", fill = "Candidate") +
  ggtitle("Ward 11 2017 Election Results By Community - In person votes only")
gWard11Col
```


Next steps: Take a look at those communities and overlay demographics maybe add Mayor bars.
compare to previous election, compare to provincial and federal elections, compare to special voters and advance voters, demographics from federal census?